The Name of the Title Is Hope

### 2024-12-30 paper.tex
my latex doesn't deal well with spaces in cite commands

\section{Background}
\subsection{The GPU Programming Environment(Model?)}
Contemporary GPUs are hierarchically organized, massively parallel processors designed to prioritize throughput over latency. Due to their specialized nature, GPUs function as co-processors to the CPU, requiring GPU programs—known as \emph{kernels} or \emph{shaders}—to be \emph{launched} or \emph{dispatched} heterogeneously by the CPU. Upon launch, a kernel is replicated across independent execution contexts, called threads. These threads are first organized into smaller units known as \emph{subgroups}\footnote{Also referred to as \emph{Warps}, \emph{Waves}, or \emph{SIMD-Groups}.}, which are then grouped into larger units called \emph{workgroups}\footnote{Also referred to as \emph{Cross-Thread-Arrays}, \emph{Thread-Blocks}, \emph{Blocks}, \emph{Thread-Groups}, or \emph{Groups}.}. This thread hierarchy mirrors the GPU memory hierarchy: individual threads access private, high-speed registers; threads within a workgroup share low-latency shared memory\footnote{Also referred to as \emph{Local Data Storage}, \emph{Group-Shared Memory}, (Check metal terminology).}; and all threads have access to high-bandwidth but higher-latency global memory.

@@

GPUs exhibit both single-instruction, multiple-thread (SIMT) and single-program, multiple-data (SPMD) behaviors. A multiprocessor schedules and executes threads by subgroup. While it attempts to execute the same instruction across all threads in the subgroup, SIMT, threads within the same subgroup are free to follow their own data/context-dependent execution paths, SPMD. However, this flexibility comes at the cost of subgroup divergence. When threads in a subgroup diverge due to conditional branching or other control flow, each unique path is executed sequentially, but at the full execution width, effectively rendering the operation serial and reducing parallel efficiency.\footnote{The scheduling and execution width of the underlying hardware does not necessarily align with the subgroup width.}

Most relevant to this work is the scheduling behavior of the GPU. Scheduling is divided into a workgroup-level scheduler\footnote{For example, NVIDIA's \emph{GigaThread Engine} or AMD's \emph{Asynchronous Compute Engines}}, which manages kernel launches and maps workgroups to multiprocessors, and the aforementioned subgroup scheduler, responsible for selecting which of the currently resident subgroups for execution. Although contemporary GPU hardware supports workgroup preemption to prioritize real-time tasks or manage multi-process workloads~\cite{}(NVIDIA Volta,AMD RDNA), this capability does not extend to the GPU programming model. Once a workgroup begins execution on a multiprocessor, it must run to completion and cannot context switch. At the subgroup scheduling level, \emph{fairness}—how evenly execution resources are distributed among threads—and \emph{progress guarantees}—ensuring that all threads or subgroups eventually make forward progress—are the most salient issues. As we will discuss in the following section, these factors have profound implications for kernel performance and correctness.

@@

\subsubsection{Scan-then-Propagate}
Introduced by Sengupta et al.~\cite{10.5555/1280094.1280110}, the \emph{Scan-then-Propagate} ~\cite{GPUGems3, Sengupta2008} strategy is an intuitive extension of the intra-workgroup scan, comprising three phases: 
\begin{enumerate} 
  \item \textbf{Intra-Workgroup (Local) Scan:} Each workgroup performs an inclusive scan on its assigned work-tile partition of the input data. The results, including both the local scan output and the reduction of each workgroup, are written back to global memory.
  \item \textbf{Spine Scan:} The reductions from all workgroups, collectively referred to as \emph{the spine}, are gathered and scanned to compute the global prefix of the reductions, resolving the dependencies between workgroups; we call the result of this phase the \emph{root}. 
  \item \textbf{Propagation:} The root is propagated back to each workgroup and combined with the local scan results to compute the final scan. 
\end{enumerate}
Without an inter-workgroup barrier primitive, each phase is implemented as a separate kernel launch to ensure a consistent view of the reductions. During the scan operation, each element is read from and written to global memory once during the local scan and again during the propagation phase, contributing a total of $4n$ global memory operations. Although data movement is dominated by the first and third phase, Sengupta et al.'s implementation handles arbitrarily large inputs using recursion. For an input of size $n$ and a work tile of size $t$, this results in a recursive depth of $\lceil \log_t n \rceil$. At each recursive step $k$, $n/t^k$ tiles are processed, leading to a total of $\sum_{k=1}^{\lceil \log_t n \rceil} n/t^k \approx (n - 1)/(t - 1)$ work tiles over all steps. Since each tile moves $4t$ data ($2t$ local scan, $2t$ propagation), the total data movement is $O\left(\frac{4t(n - 1)}{t - 1}\right) = O(4n)$.

\subsubsection{Reduce-then-Scan}
Due to the extreme latency of global memory access relative to compute, computationally light algorithms like \emph{scan} can benefit from redundant calculation if it reduces memory access. This insight led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}~\cite{Merrill Grimshaw, Ha and Han, Dotsenko, Breitbart}: by writing only the work tile reduction in the first phase and performing a redundant scan in the third phase, $n$ global data movement can be eliminated. The \emph{Reduce-Then-Scan} strategy operates in three phases:
\begin{enumerate}
  \item \textbf{Work Tile Reduction:} Each workgroup performs a pure reduction on its work tile partition, posting the result into global memory. 
  \item \textbf{Spine Scan:} The reductions from all workgroups are gathered and scanned to produce the root.
  \item \textbf{Intra-Workgroup Scan and Propagation:} Each workgroup performs the local scan on the work tile, incorporating the root as it writes to global memory.
\end{enumerate}
Merrill and Grimshaw~\cite{} introduced a workgroup-level \emph{raking} technique that improved the spine phase of \emph{Reduce-Then-Scan}. Instead of handling arbitrary inputs recursively, this approach determines the total workgroup occupancy $o$ and partitions the input into non-overlapping blocks, processed serially by each workgroup. By limiting the recursive depth of the global scan to two, the global memory movement for the spine scan is reduced from $O(n/t)$ to $O(3o)$. Consequently, kernel launches are limited to three, and total global memory movement is reduced to $O(3n + 3o) = O(3n)$. This approach was further refined by Ha and Han~\cite{}, who modified the upsweep behavior of the first workgroup from reduction to scan and eliminated the redundant upsweep of the last workgroup, thereby saving the equivalent of one workgroup's computation.

\subsubsection{Chained-Scan}
Continuing the trend of global data movement optimization, the most recent work on global scan has been the \emph{Chained-Scan} architecture. First introduced by Yan et al. in \emph{StreamScan}, the key innovation of GPU chained scanning lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level. Instead of using kernel launches as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups. This shifts the problem from explicit synchronization to implicit synchronization via the order in which workgroups access global data. Each workgroup processes its assigned tile independently, waiting only for the result of the preceding reductions before continuing. This structure ensures that global memory traffic is minimized, as each element is read and written exactly once, achieving the theoretical minimum of $2n$.

@@

For instance, consider a scan operation requiring two workgroups running on a GPU with a single multiprocessor. Although the multiprocessor can host both workgroups, it has only one scheduling unit, which lacks FPG. In this setup, both workgroups launch and acquire their respective work tiles. The workgroup responsible for \emph{Work Tile 1} finishes first and begins waiting for the reduction of \emph{Work Tile 0}. However, because the scheduler lacks FPG, it may continue scheduling \emph{Work Tile 1}, leaving \emph{Work Tile 0} starved and unable to complete its reduction. As a result, \emph{Work Tile 1} spins indefinitely, eventually triggering \emph{timeout detection and recovery} (TDR) and crashing the host program.

\subsection{Why is reliance on Forward Progress Guarantees a Portability Problem?}
Although NVIDIA formalized FPG down to the intra-subgroup level beginning with the Volta architecture---and FPG was likely already present at the workgroup level on at least NVIDIA's Fermi and AMD's TeraScale2 architectures~\cite{}---it remains absent on some contemporary hardware, most notably certain ARM-based chips, including Apple's M-series GPUs~\cite{10.1145/3485508}. Our testing shows that, at best, attempting to run \emph{Chained-Scan} without FPG results in mega-bad\footnote{Need to gather data on the distribution of M1 Pro time on Decoupled Lookback instead of the average time. It's also unclear why the test did not trigger Metal's timeout device recovery}, and at worst, risks the afformentioned TDR and subsequent program crash. This extreme slowdown highlights the inherent fragility of \emph{Chained-Scan} without FPG.

Exacerbating this issue is the fact that no major graphics API currently provides developers with a mechanism to query FPG support\footnote{As of writing this, 12/29/2024, we are unaware of any major graphics API---Vulkan, D3D12, or Metal---that provide developers with an FPG capability query}. This lack of transparency forces developers to design algorithms without clear knowledge of whether critical guarantees are available on the target hardware. In environments like shading languages, where cross-platform portability is required, this uncertainty compels developers to revert to older, less efficient strategies such as \emph{Reduce-Then-Scan}. However, as discussed earlier, \emph{Reduce-Then-Scan} carries its own set of drawbacks, including increased global memory traffic ($O(3n)$) and additional kernel launches, which result in reduced throughput and higher latency. While these tradeoffs were once deemed acceptable, they stand in stark contrast to the efficiency and simplicity of \emph{Chained-Scan} when implemented on hardware with reliable FPG. Yet, in the absence of guaranteed forward progress, developers have little choice but to sacrifice performance for broader compatibility.

@@

\emph{ScallarFallback} suffers from two primary issues that impact its efficiency. First, the fallback reduction is performed serially by a single thread, progressing only when the workgroup engages in a waiting spin.This effectively takes $t$ spins to complete the fallback of the deadlocking tile. Moreover, because only a single thread is responsible for loading elements, the cache line is severely underutilized, further degrading performance. Second, once the fallback is complete, the workgroup does not post its result to device memory. Consequently, every workgroup deadlocked by that tile must redundantly execute the fallback, wasting computational resources. While \emph{ScallarFallback} was able to achieve speed of light on hardware supporting FPG and executed correctly on hardware without FPG, its performance on the latter was inferior to that of \emph{Reduce-Then-Scan} .

\section{Design}
\subsection{Goals}
This work is guided by two main goals. First, portability: we want to develop a scan implementation that retains the benefits of the \emph{Chained-Scan} architecture but is also suitable for a diverse range of hardware vendors and achitectures, including those without FPG. Second, performance: we aim for near speed-of-light execution to the greatest extent possible allowed by the underlying hardware and programming model. To ground these objectives, we select the WebGPU shading language (WGSL) and the WebGPU standard as our target environment. WGSL is a meta-level shading language which is translated and compiled as necessary for backend graphics APIs---D3D12, Vulkan, and Metal---and as such, WGSL implementations are bound by the minimum capability across all backends. Because WGSL must operate within this limited capability space, it inherently embodies the challenge of portability.

@@

\subsection{Intra-Workgroup Implementation}
Our intra-workgroup implementation is most similar to the \emph{CUDPP} scan kernel ~\cite{}, with adaptations to enhance portability. It consists of three phases:
\begin{enumerate}
  \item
\end{enumerate}

@@

>>>>

\section{Background}
\subsection{The GPU Programming Environment(Model?)}
Contemporary GPUs are hierarchically organized, massively parallel processors designed to prioritize throughput over latency. Due to their specialized nature, GPUs function as co-processors to the CPU, requiring GPU programs—known as \emph{kernels} or \emph{shaders}—to be \emph{launched} or \emph{dispatched} heterogeneously by the CPU\@. Upon launch, a kernel is replicated across independent execution contexts, called threads. These threads are first organized into smaller units known as \emph{subgroups}\footnote{Also referred to as \emph{Warps}, \emph{Waves}, or \emph{SIMD-Groups}.}, which are then grouped into larger units called \emph{workgroups}\footnote{Also referred to as \emph{Cross-Thread-Arrays}, \emph{Thread-Blocks}, \emph{Blocks}, \emph{Thread-Groups}, or \emph{Groups}.}. This thread hierarchy mirrors the GPU memory hierarchy: individual threads access private, high-speed registers; threads within a workgroup share low-latency shared memory\footnote{Also referred to as \emph{Local Data Storage}, \emph{Group-Shared Memory}, (Check metal terminology).}; and all threads have access to high-bandwidth but higher-latency global memory.

@@

GPUs exhibit both single-instruction, multiple-thread (SIMT) and single-program, multiple-data (SPMD) behaviors. A multiprocessor schedules and executes threads by subgroup. While it attempts to execute the same instruction across all threads in the subgroup, SIMT, threads within the same subgroup are free to follow their own data/context-dependent execution paths, SPMD\@. However, this flexibility comes at the cost of subgroup divergence. When threads in a subgroup diverge due to conditional branching or other control flow, each unique path is executed sequentially, but at the full execution width, effectively rendering the operation serial and reducing parallel efficiency.\footnote{The scheduling and execution width of the underlying hardware does not necessarily align with the subgroup width.}

Most relevant to this work is the scheduling behavior of the GPU\@. Scheduling is divided into a workgroup-level scheduler\footnote{For example, NVIDIA's \emph{GigaThread Engine} or AMD's \emph{Asynchronous Compute Engines}}, which manages kernel launches and maps workgroups to multiprocessors, and the aforementioned subgroup scheduler, responsible for selecting which of the currently resident subgroups for execution. Although contemporary GPU hardware supports workgroup preemption to prioritize real-time tasks or manage multi-process workloads~\cite{}(NVIDIA Volta,AMD RDNA), this capability does not extend to the GPU programming model. Once a workgroup begins execution on a multiprocessor, it must run to completion and cannot context switch. At the subgroup scheduling level, \emph{fairness}—how evenly execution resources are distributed among threads—and \emph{progress guarantees}—ensuring that all threads or subgroups eventually make forward progress—are the most salient issues. As we will discuss in the following section, these factors have profound implications for kernel performance and correctness.

@@

\subsubsection{Scan-then-Propagate}
Introduced by Sengupta et al.~\cite{10.5555/1280094.1280110}, the \emph{Scan-then-Propagate}~\cite{GPUGems3, Sengupta2008} strategy is an intuitive extension of the intra-workgroup scan, comprising three phases:
\begin{enumerate}
  \item \textbf{Intra-Workgroup (Local) Scan:} Each workgroup performs an inclusive scan on its assigned work-tile partition of the input data. The results, including both the local scan output and the reduction of each workgroup, are written back to global memory.
  \item \textbf{Spine Scan:} The reductions from all workgroups, collectively referred to as \emph{the spine}, are gathered and scanned to compute the global prefix of the reductions, resolving the dependencies between workgroups; we call the result of this phase the \emph{root}.
  \item \textbf{Propagation:} The root is propagated back to each workgroup and combined with the local scan results to compute the final scan.
\end{enumerate}
Without an inter-workgroup barrier primitive, each phase is implemented as a separate kernel launch to ensure a consistent view of the reductions. During the scan operation, each element is read from and written to global memory once during the local scan and again during the propagation phase, contributing a total of $4n$ global memory operations. Although data movement is dominated by the first and third phase, Sengupta et al.'s implementation handles arbitrarily large inputs using recursion. For an input of size $n$ and a work tile of size $t$, this results in a recursive depth of $\lceil \log_t n \rceil$. At each recursive step $k$, $n/t^k$ tiles are processed, leading to a total of $\sum_{k=1}^{\lceil \log_t n \rceil} n/t^k \approx (n - 1)/(t - 1)$ work tiles over all steps. Since each tile moves $4t$ data ($2t$ local scan, $2t$ propagation), the total data movement is $O\left(\frac{4t(n - 1)}{t - 1}\right) = O(4n)$.

\subsubsection{Reduce-then-Scan}
Due to the extreme latency of global memory access relative to compute, computationally light algorithms like \emph{scan} can benefit from redundant calculation if it reduces memory access. This insight led to a refinement of the \emph{Scan-Then-Propagate} strategy called \emph{Reduce-Then-Scan}~\cite{Merrill-Grimshaw, Ha-and-Han, Dotsenko, Breitbart}: by writing only the work tile reduction in the first phase and performing a redundant scan in the third phase, $n$ global data movement can be eliminated. The \emph{Reduce-Then-Scan} strategy operates in three phases:
\begin{enumerate}
  \item \textbf{Work Tile Reduction:} Each workgroup performs a pure reduction on its work tile partition, posting the result into global memory.
  \item \textbf{Spine Scan:} The reductions from all workgroups are gathered and scanned to produce the root.
  \item \textbf{Intra-Workgroup Scan and Propagation:} Each workgroup performs the local scan on the work tile, incorporating the root as it writes to global memory.
\end{enumerate}
Merrill and Grimshaw~\cite{} introduced a workgroup-level \emph{raking} technique that improved the spine phase of \emph{Reduce-Then-Scan}. Instead of handling arbitrary inputs recursively, this approach determines the total workgroup occupancy $o$ and partitions the input into non-overlapping blocks, processed serially by each workgroup. By limiting the recursive depth of the global scan to two, the global memory movement for the spine scan is reduced from $O(n/t)$ to $O(3o)$. Consequently, kernel launches are limited to three, and total global memory movement is reduced to $O(3n + 3o) = O(3n)$. This approach was further refined by Ha and Han~\cite{}, who modified the upsweep behavior of the first workgroup from reduction to scan and eliminated the redundant upsweep of the last workgroup, thereby saving the equivalent of one workgroup's computation.

\subsubsection{Chained-Scan}
Continuing the trend of global data movement optimization, the most recent work on global scan has been the \emph{Chained-Scan} architecture. First introduced by Yan et al.\ in \emph{StreamScan}, the key innovation of GPU chained scanning lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level. Instead of using kernel launches as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups. This shifts the problem from explicit synchronization to implicit synchronization via the order in which workgroups access global data. Each workgroup processes its assigned tile independently, waiting only for the result of the preceding reductions before continuing. This structure ensures that global memory traffic is minimized, as each element is read and written exactly once, achieving the theoretical minimum of $2n$.

@@

For instance, consider a scan operation requiring two workgroups running on a GPU with a single multiprocessor. Although the multiprocessor can host both workgroups, it has only one scheduling unit, which lacks FPG\@. In this setup, both workgroups launch and acquire their respective work tiles. The workgroup responsible for \emph{Work Tile 1} finishes first and begins waiting for the reduction of \emph{Work Tile 0}. However, because the scheduler lacks FPG, it may continue scheduling \emph{Work Tile 1}, leaving \emph{Work Tile 0} starved and unable to complete its reduction. As a result, \emph{Work Tile 1} spins indefinitely, eventually triggering \emph{timeout detection and recovery} (TDR) and crashing the host program.

\subsection{Why is reliance on Forward Progress Guarantees a Portability Problem?}
Although NVIDIA formalized FPG down to the intra-subgroup level beginning with the Volta architecture---and FPG was likely already present at the workgroup level on at least NVIDIA's Fermi and AMD's TeraScale2 architectures~\cite{}---it remains absent on some contemporary hardware, most notably certain ARM-based chips, including Apple's M-series GPUs~\cite{10.1145/3485508}. Our testing shows that, at best, attempting to run \emph{Chained-Scan} without FPG results in mega-bad\footnote{Need to gather data on the distribution of M1 Pro time on Decoupled Lookback instead of the average time. It's also unclear why the test did not trigger Metal's timeout device recovery}, and at worst, risks the afformentioned TDR and subsequent program crash. This extreme slowdown highlights the inherent fragility of \emph{Chained-Scan} without FPG\@.

Exacerbating this issue is the fact that no major graphics API currently provides developers with a mechanism to query FPG support\footnote{As of writing this, 12/29/2024, we are unaware of any major graphics API---Vulkan, D3D12, or Metal---that provide developers with an FPG capability query}. This lack of transparency forces developers to design algorithms without clear knowledge of whether critical guarantees are available on the target hardware. In environments like shading languages, where cross-platform portability is required, this uncertainty compels developers to revert to older, less efficient strategies such as \emph{Reduce-Then-Scan}. However, as discussed earlier, \emph{Reduce-Then-Scan} carries its own set of drawbacks, including increased global memory traffic ($O(3n)$) and additional kernel launches, which result in reduced throughput and higher latency. While these tradeoffs were once deemed acceptable, they stand in stark contrast to the efficiency and simplicity of \emph{Chained-Scan} when implemented on hardware with reliable FPG\@. Yet, in the absence of guaranteed forward progress, developers have little choice but to sacrifice performance for broader compatibility.

@@

\emph{ScallarFallback} suffers from two primary issues that impact its efficiency. First, the fallback reduction is performed serially by a single thread, progressing only when the workgroup engages in a waiting spin.This effectively takes $t$ spins to complete the fallback of the deadlocking tile. Moreover, because only a single thread is responsible for loading elements, the cache line is severely underutilized, further degrading performance. Second, once the fallback is complete, the workgroup does not post its result to device memory. Consequently, every workgroup deadlocked by that tile must redundantly execute the fallback, wasting computational resources. While \emph{ScallarFallback} was able to achieve speed of light on hardware supporting FPG and executed correctly on hardware without FPG, its performance on the latter was inferior to that of \emph{Reduce-Then-Scan}.

\section{Design}
\subsection{Goals}
This work is guided by two main goals. First, portability: we want to develop a scan implementation that retains the benefits of the \emph{Chained-Scan} architecture but is also suitable for a diverse range of hardware vendors and achitectures, including those without FPG\@. Second, performance: we aim for near speed-of-light execution to the greatest extent possible allowed by the underlying hardware and programming model. To ground these objectives, we select the WebGPU shading language (WGSL) and the WebGPU standard as our target environment. WGSL is a meta-level shading language which is translated and compiled as necessary for backend graphics APIs---D3D12, Vulkan, and Metal---and as such, WGSL implementations are bound by the minimum capability across all backends. Because WGSL must operate within this limited capability space, it inherently embodies the challenge of portability.

@@

\subsection{Intra-Workgroup Implementation}
Our intra-workgroup implementation is most similar to the \emph{CUDPP} scan kernel~\cite{}, with adaptations to enhance portability. It consists of three phases:
\begin{enumerate}
  \item
\end{enumerate}

@@

%%% Local Variables:
%%% mode: LaTeX
%%% TeX-master: t
%%% End:
